Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-02 manuscript/manuscript.tex
update manuscript

% Created 2019-10-19 Sat 12:19
% Intended LaTeX compiler: pdflatex
\documentclass[twocolumn, 10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=raggedright,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$,Christine Lonjou$^{1,2,3}$,Fabienne Lesueur$^{1,2,3}$,GENESIS
  investigators,Dominique Stoppa-Lyonnet$^{4,5,6}$,Nadine Andrieu$^{1,2,3}$,Chloé-Agathe Azencott$^{3,1,2}$}
\affiliation{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France; \\
  $^{2}$INSERM, U900, F-75005 Paris, France; \\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^4$Service de Génétique, Institut Curie, F-75005 Paris, France;
  $^5$INSERM, U830, F-75005 Paris, France;
  $^6$Université Paris Descartes;

}
\hypersetup{
 pdfauthor={Héctor Climente-González, Chloé-Agathe Azencott},
 pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 26.2 (Org mode 9.2.3)}, 
 pdflang={English}}
\begin{document}

\onecolumn
\MakeTitle
\begin{abstract}

Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between a gene and a phenotype, and the biological context of each gene, represented as a network. In this work we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to a phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer.

\end{abstract}

\section{Introduction}
\label{sec:orgbee581e}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The omnigenic model \cite{boyle_expanded_2017} explains this by the dense functional inter-relatedness between genes, influencing each other's behavior, which allows alterations in most genes to impact the ``core'' genes involved in a disease. A comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships between them are the edges that link them. These functional relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their gene products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent biomolecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011} \cite{pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, guilt-by-association strategies have proved effective at identifying disease genes \cite{huang_systematic_2018}.

@@

In this work, we analyze the effectiveness of these six methods to discover new biomarkers on GWAS data. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After following a classical GWAS approach, we use these network-based methods to recover additional familial breast cancer biomarkers. Some of them are known, while others are specific to this dataset. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

\section{Methods}
\label{sec:org9e7ca64}
\subsection{GENESIS}
\label{sec:org920eb00}

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for BRCA1 and BRCA2 pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of the corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

\subsection{Preprocessing and quality control}
\label{sec:orgd26c48b}

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless 0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes, and an additional 28 samples with TODO. The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could create confounding associations. A PCA revealed no differential population structure between cases and controls (Supplementary Figure \ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\) = 1.05) either, further confirming the absence of confounding population structure.

\subsection{High-score subnetwork search algorithms}
\label{sec:org6ced6c9}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1df \(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that could be mapped to an HGNC symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 mapped to at least one of these genes.

We use such mapping to compare the outputs of methods who produce SNP- to those that produce gene-lists, and vice versa. In the former case, we consider any gene that can be mapped to any of the selected SNPs as selected as well. In the latter, we consider all the SNPs that can be mapped to that gene as selected by the method.

@@

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected through an edge, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes. Lastly, the \emph{Laplacian matrix} $L_G$ is defined as $L_G = D_G - W_G$.

@@

\begin{table}[htbp]
  \begin{threeparttable}
    \caption{\label{tab:method_comparison}
      Summary of the differences between the studied algorithms.}
    \centering
    \begin{tabular}{l|lllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Modular & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Modular & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Modular & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      /Field/: field in which the algorithm was developed. /Nodes/, the type of network, either gene (protein-protein interaction network usually) or a SNP network. /Exhaustive/: whether all the possible solutions given the selected hyperparameters are explored. /Solution/: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. /Components/: number of connected subnetworks in the solution. /Input/: genotype data or GWAS summary statistics. /Scoring/: how SNP/gene P-values are transformed into node scores.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Beyond the assumption that genes that contribute to the same function will be nearby in the protein-protein interaction network (PPIN), they might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). That is not the only choice to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, what is the best network to use, etc. In consequence, multiple solutions have been proposed. In this article, we examine six of them: five that explore the protein-protein interaction network, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table \ref{tab:method_comparison}.

@@

\begin{enumerate}
\item Select a seed node.
\item Compute Stouffer's Z-score Z\textsubscript{m} for the current subgraph as

\begin{equation*} 
Z_m = \frac{\sum z_i}{\sqrt{k}}
\end{equation*}

where \emph{k} is the number of genes in the subgraph, z\textsubscript{i} = \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{i}), and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes i.e. nodes at distance \(\le\) \emph{d}. We set d = 2.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than Z\textsubscript{m} \texttimes{} (1 + r). In our experiments, we set r = 0.1.
\item Repeat 2-4 until no increment Z\textsubscript{m} \texttimes{} (1 + r) is possible.
\end{enumerate}

@@

where Z\textsubscript{m}(\(\pi\)) represent a vector containing 100000 random subsets of the same number of genes.

@@

were p(v) = w(v) - w' is the \emph{profit} of adding a node, c(e) = w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. heinz implements the algorithm from \cite{ljubic_algorithmic_2006}, which in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz, available on Bioconductor \cite{beisser_bionet:_2010,heinz}.

@@

where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. We scored the nodes as in \cite{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. To obtain densely connected subnetworks, HotNet2 prunes E, only preserving edges such that w(E) \textgreater{} \(\delta\). Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, and are robust \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given two hyperparameters, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. We used the version on SConES implemented in the R package martini \cite{martini}. As in \cite{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. Note that the solution to the above problem can consist of several connected subnetworks which are disconnected from each other. In this case, the selected hyperparameters were \(\eta\) = 3.51, \(\lambda\) = 210.29 for SConES GS; \(\eta\) = 3.51, \(\lambda\) = 97.61 for SConES GM; and \(\eta\) = 3.51, \(\lambda\) = 45.31 for SConES GI.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the most densely connected gene subnetwork that is most strongly associated to the phenotype \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the Laplacian matrix rather than the adjacency matrix (Section \ref{methods:notation}), to quantify solutions containing many edges.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, by replacing the Laplacian by the adjacency matrix, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\) = \(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is strongly interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

\subsubsection{Gene-gene network}
\label{methods:gene_network}
Out of the six methods tested, five use a gene-gene interaction network (Section \ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), for practical reasons we focused on a PPIN, as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to it. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless specified otherwise, we used only interactions coming from high-throughput experiments to avoid biasing the topology of the network by well-studied genes with more known interactions on average. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of HGNC symbols. The scoring function for the nodes changed from method to method (Section \ref{methods:methods}).

>>>>

\documentclass[twocolumn, 10pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=raggedright,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, GENESIS study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$, Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\affiliation{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes;\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS investigators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

\onecolumn
\MakeTitle

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. It does so by jointly considering the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six network methods to discover new biomarkers for breast cancer susceptibility by searching subnetworks highly associated to this phenotype. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value = $7.8 \times 10^{-5}$ ) and more central than average. Additionally, it includes subnetworks of mechanisms related to cancer, like protein folding (\emph{HSPA1A}, \emph{HSPA1B}, and \emph{HSPA1L}) or mitochondrial ribosomes (\emph{MRPS30}, \emph{MRPS31}, \emph{MRPS18B}). We also observed a general dysregulation in the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. By trading statistical astringency for biological meaningfulness, most network methods get more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. 
\end{abstract}

\section{Introduction}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, the experimental setting also presents intrinsic challenges. Some of them stem from the high-dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with large and moderate effects can be detected. And it is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically Bonferroni. However, Bonferroni is known to be overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the ``core'' genes directly involved in a disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

In this regard many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and between them are the edges that link them. These functional relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011} \cite{pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, guilt-by-association strategies have proved effective at identifying disease genes \cite{huang_systematic_2018}.

@@

In this work, we analyze the effectiveness of these six methods to discover new biomarkers on GWAS data. We focus on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study of familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network-based methods to recover additional breast cancer biomarkers. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus network of predisposition to familial breast cancer.

\section{Methods}
\subsection{GENESIS}

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for \emph{BRCA1} and \emph{BRCA2} pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of the corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

\subsection{Preprocessing and quality control}

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy - Weinberg equilibrium in controls (P-value \textless 0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After control for relatedness, 17 additional samples were removed (6 for sample identity error, 6 false ``friend/control'' having family link with other samples, 3 ``friend/control'' having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with CLIS, 1 with \emph{BRCA1/2} mutation detected in the family, 1 with molecular diagnosis not received). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could create confounding associations. A PCA revealed no visual differential population structure between cases and controls (Supplementary Figure \ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\) = 1.05) either, further confirming the absence of confounding population structure.

\subsection{High-score subnetwork search algorithms}
\subsubsection{SNP and gene association}
\label{methods:node_score}
To measure association between a genotype and the phenotype, we performed a per-SNP 1~d.f.~\(\chi\)\textsuperscript{2} allelic test using PLINK v1.90 \cite{chang_second-generation_2015}. Then, we used VEGAS2v2 to compute the gene-level association score \cite{mishra_vegas2:_2015} from the SNP P-values. In order to map SNPs to genes we used their overlap on the genome: all SNPs located within the boundaries of a gene, \textpm 50 kb, were mapped to that gene. To compute the gene association we used the 10\% of SNPs linked to the gene with lowest P-values. We used the 62\,193 genes described in GENCODE 31 \cite{frankish_gencode_2019}, although only 54\,612 could be mapped to at least one SNP. Out of those, we focused exclusively on the 32\,767 that had a gene symbol. Out of the SNPs 197\,083 remaining after quality control, 164\,037 were mapped to at least one of these genes.

We use such mapping to compare the outputs of methods that produce SNP- to those that produce gene-lists, and vice versa. In the former case, we consider any gene that can be mapped to any of the selected SNPs as selected as well. In the latter, we consider all the SNPs that can be mapped to that gene as selected by the method.

@@

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes. Lastly, the \emph{Laplacian matrix} $L_G$ is defined as $L_G = D_G - W_G$.

@@

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Components & Input & Scoring & Reference\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \textbf{\emph{Field}}: field in which the algorithm was developed.\\
      \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network.\\
      \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored.\\
      \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected.\\
      \textbf{\emph{Components}}: number of connected subnetworks in the solution.\\
      \textbf{\emph{Input}}: genotype data or GWAS summary statistics.\\
      \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores.\\
      \textbf{\emph{Reference}}: original publication featuring the algorithm.
    \end{tablenotes}
  \end{threeparttable}
\end{table}

Beyond the assumption that genes that contribute to the same function will be nearby in the protein-protein interaction network (PPIN), they might be topologically related to each other in diverse ways (densely interconnected modules, nodes around a hub, a path, etc.). That is not the only choice to make: how to score the nodes, whether the affected mechanisms form a single connected component or several, how to frame the problem in a computationally efficient fashion, what is the best network to use, etc. In consequence, multiple solutions have been proposed. In this article, we examine six of them: five that explore the PPIN, and one which explores SNP networks. We selected methods that were open source, had an implementation available, and an accessible documentation. Their main differences are summarized in Table \ref{tab:method_comparison}.

@@

\begin{enumerate}
\item Select a seed node.
\item Compute Stouffer's Z-score Z\textsubscript{m} for the current subgraph $S$ as

\begin{equation*} 
Z_m = \frac{\sum_{i \in S} z_i}{\sqrt{k}}
\end{equation*}

where \emph{k} is the number of genes in the subgraph; z\textsubscript{i} is the score of gene $i$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{i}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes i.e. nodes at distance \(\le\) \emph{d}. We set d = 2.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than Z\textsubscript{m} \texttimes{} (1 + r). In our experiments, we set r = 0.1.
\item Repeat 2-4 until no increment Z\textsubscript{m} \texttimes{} (1 + r) is possible.
\end{enumerate}

@@

where Z\textsubscript{m}(\(\pi\)) represent a vector containing 100\,000 random subsets of the same number of genes.

@@

were p(v) = w(v) - w' is the \emph{profit} of adding a node, c(e) = w' is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$. All three are positive quantities. heinz implements the algorithm from \citet{ljubic_algorithmic_2006}, which in practice is often fast and optimal, neither is guaranteed. We used BioNet's implementation of heinz, available on Bioconductor \cite{beisser_bionet:_2010,heinz}.

@@

where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. We scored the nodes as in \citet{nakka_gene_2016}, assigning a score of 0 for the genes with low probability of being associated to the disease, and -log\textsubscript{10}(P-value) to those likely to be. In this dataset, the threshold separating both was a P-value of 0.125, which was obtained using a local FDR approach \cite{scheid_twilight;_2005}. To obtain densely connected subnetworks, HotNet2 prunes E, only preserving edges such that w(E) \textgreater{} \(\delta\). Lastly, HotNet2 evaluates the statistical significance of the subnetworks by comparing their size to the size of networks obtained by permuting the node scores. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, and are robust \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are hyperparameters that control the sparsity and the connectivity of the model. Given two hyperparameters, the aforementioned problem has a unique solution, that SConES finds using a graph min-cut procedure. We used the version on SConES implemented in the R package martini \cite{martini}. As in \citet{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. Note that the solution to the above problem can consist of several connected subnetworks which are disconnected from each other. In this case, the selected hyperparameters were \(\eta\) = 3.51, \(\lambda\) = 210.29 for SConES GS; \(\eta\) = 3.51, \(\lambda\) = 97.61 for SConES GM; and \(\eta\) = 3.51, \(\lambda\) = 45.31 for SConES GI.
\end{description}

\begin{description}
\item[{SigMod}] SigMod aims at identifying the most densely connected gene subnetwork that is most strongly associated to the phenotype \cite{liu_sigmod:_2017}. It addresses an optimization problem similar to that of SConES (Equation \ref{eq:scones}), but using the adjacency matrix rather than the Laplacian matrix (Section \ref{methods:notation}), to quantify solutions containing many edges.

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, by replacing the adjacency by the Laplacian matrix, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\) = \(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is strongly interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

\begin{description}
\item[{Consensus}] In addition, we built a consensus network by retaining the nodes that were selected by at least two of the six methods (using SConES GI for SConES).
\end{description}

\subsubsection{Gene-gene network}
\label{methods:gene_network}
Out of the six methods tested, five use a gene-gene interaction network (Section \ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), for practical reasons we focused on a PPIN, as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to it. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless specified otherwise, we used only interactions coming from high-throughput experiments to avoid biasing the topology of the network by well-studied genes with more known interactions on average. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section \ref{methods:methods}).
